\textbf{Adaptation of the PHQ-4 Short Screening for Depression and Anxiety to increase its Sensitivity to Subclinical Variability}

### 2023-01-08 paper/manuscript.Rmd
formatting

The Patient Health Questionnaire-4 (PHQ-4) is an ultra brief measurement of core signs of depression and anxiety [@kroenke2009ultra]. It consists of two items for depression [PHQ–2, @kroenke2003patient] and anxiety [GAD–2, @kroenke2007anxiety], each corresponding to DSM-V Diagnostic Criterion A symptoms for major depressive disorder (MDD) and generalized anxiety disorder (GAD). It has been validated across many languages and populations [@christodoulaki2022validation, @mendoza2022factor, @materu2020psychometric], becoming one of the most popular screening instruments for depression and anxiety [@maurer2018depression].

While the scale has been validated and used in the general population [@lowe20104, @hajek2020prevalence], its initial purpose was to reliably discriminate and identify potential MDD/GAD patients. This discriminative goal materializes in the scale's design and the existence of categorical cut-offs, which does not necessary entail a strong sensitivity to milder mood alterations. In particular, the gap between the two lowest possible answers, "Not at all" and "Several days", is quite large and leaves out the possibility of more subtle occurrences. While this is not necessarily an issue in clinical and diagnostic contexts, it might lead to a sub-optimal discrimination of affective levels on the lower end of the spectrum, important for instance in the context of subclinical variability quantifications. The goal of this study is hence to enhance, with minimal changes to the original scale, the sensitivity to mild mood level inflections.

>>>>

The Patient Health Questionnaire-4 (PHQ-4) is an ultra brief measurement of core signs of depression and anxiety [@kroenke2009ultra]. It consists of two items for depression [PHQ–2, @kroenke2003patient] and anxiety [GAD–2, @kroenke2007anxiety], each corresponding to DSM-V Diagnostic Criterion A symptoms for major depressive disorder (MDD) and generalized anxiety disorder (GAD). It has been validated across many languages and populations [@christodoulaki2022validation; @mendoza2022factor; @materu2020psychometric], becoming one of the most popular screening instruments for depression and anxiety [@maurer2018depression].

While the scale has been validated and used in the general population [@lowe20104; @hajek2020prevalence], its initial purpose was to reliably discriminate and identify potential MDD/GAD patients. This discriminative goal materializes in the scale's design and the existence of categorical cut-offs, which does not necessary entail a strong sensitivity to milder mood alterations. In particular, the gap between the two lowest possible answers, "Not at all" and "Several days", is quite large and leaves out the possibility of more subtle occurrences. While this is not necessarily an issue in clinical and diagnostic contexts, it might lead to a sub-optimal discrimination of affective levels on the lower end of the spectrum, important for instance in the context of subclinical variability quantifications. The goal of this study is hence to enhance, with minimal changes to the original scale, the sensitivity to mild mood level inflections.

### 2023-01-08 paper/manuscript.tex
formatting

The Patient Health Questionnaire-4 (PHQ-4) is an ultra brief measurement of core signs of depression and anxiety (Kroenke et al., 2009). It consists of two items for depression (PHQ--2, Kroenke et al., 2003) and anxiety (GAD--2, Kroenke et al., 2007), each corresponding to DSM-V Diagnostic Criterion A symptoms for major depressive disorder (MDD) and generalized anxiety disorder (GAD). It has been validated across many languages and populations Materu et al. (2020), becoming one of the most popular screening instruments for depression and anxiety (Maurer et al., 2018).

While the scale has been validated and used in the general population Hajek \& König (2020), its initial purpose was to reliably discriminate and identify potential MDD/GAD patients. This discriminative goal materializes in the scale's design and the existence of categorical cut-offs, which does not necessary entail a strong sensitivity to milder mood alterations. In particular, the gap between the two lowest possible answers, ``Not at all'' and ``Several days'', is quite large and leaves out the possibility of more subtle occurrences. While this is not necessarily an issue in clinical and diagnostic contexts, it might lead to a sub-optimal discrimination of affective levels on the lower end of the spectrum, important for instance in the context of subclinical variability quantifications. The goal of this study is hence to enhance, with minimal changes to the original scale, the sensitivity to mild mood level inflections.

>>>>

The Patient Health Questionnaire-4 (PHQ-4) is an ultra brief measurement of core signs of depression and anxiety (Kroenke et al., 2009). It consists of two items for depression (PHQ--2, Kroenke et al., 2003) and anxiety (GAD--2, Kroenke et al., 2007), each corresponding to DSM-V Diagnostic Criterion A symptoms for major depressive disorder (MDD) and generalized anxiety disorder (GAD). It has been validated across many languages and populations (Christodoulaki et al., 2022; Materu et al., 2020; Mendoza et al., 2022), becoming one of the most popular screening instruments for depression and anxiety (Maurer et al., 2018).

While the scale has been validated and used in the general population (Hajek \& König, 2020; Löwe et al., 2010), its initial purpose was to reliably discriminate and identify potential MDD/GAD patients. This discriminative goal materializes in the scale's design and the existence of categorical cut-offs, which does not necessary entail a strong sensitivity to milder mood alterations. In particular, the gap between the two lowest possible answers, ``Not at all'' and ``Several days'', is quite large and leaves out the possibility of more subtle occurrences. While this is not necessarily an issue in clinical and diagnostic contexts, it might lead to a sub-optimal discrimination of affective levels on the lower end of the spectrum, important for instance in the context of subclinical variability quantifications. The goal of this study is hence to enhance, with minimal changes to the original scale, the sensitivity to mild mood level inflections.
